---
title: "White-faced Darter distribution in the Scottish Highlands is related to wet pine forests"
authors: "Matthew Geary & Achaz von Hardenberg
output:
  pdf_document: 
    latex_engine: xelatex
  html_notebook: default
---

---
title: White-faced darter distribution in the Scottish Highlands
author: "Matthew Geary & Achaz von Hardenberg"
date: "2018"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE)
library(raster)
<<<<<<< HEAD
library(AICcmodavg)
library(MuMIn)
=======
library(dismo)
library(unmarked)
library(rasterVis)
library(ENMeval)
>>>>>>> 16b6bf280b573eb658e8971e4d8fbacb67a81b80
```

# Main message: WHITE-FACED DARTER OCCUPANCY IS POSITIVELY RELATED TO THE PRESENCE OF WET FOREST HABITATS. THEY ARE LIKELY TO OCCUPY MORE HABITAT WITHIN THE SCOTTISH HIGHLANDS THAN IS CURRENTLY RECORDED. 

## Abstract

1) Invertebrate species are often overlooked in conservation ecology research
3) In particular, our understanding of dragonfly distributions is often geographically comprehensive but less so in ecological terms
4) White-faced darter is a lowland peatbog specialist dragonfly which has experienced population declines in Great Britain. They are thought to rely on peat-rich pool complexes within woodland but this has not yet been empirically tested
5) We used dragonfly recording data from XXX to XXX to investigate habitat preference for White-faced darter using MaxEnt across large scales in the highlands of Scotland
6) Our best models used the interaction of pine forest and wet, bog landscapes as predictors and suggested that the probability of White-faced darter occupancy increases with the local proportions of both bog and forest
7) Our models suggest that White-faced darter may be present in currently unknown sites across the highlands and their management should concentrate on maintaining sphagnum rich pool complexes and the maintenance and restoration of native forests in which these pool complexes occur




Introduction
--------------

* Dragonflies

Globally conservation research, intervention and management is biased towards high-profile taxa [REF]. Disproportionate effort is afforded to mammals and birds in particular, whereas invertebrates, which make up the majority of species present on Earth, receive relatively little attention [REF]. Cardoso _et al._ [REF] suggest that the reasons for this discrepancy can be explained by a series of potential impediments that reduce support for invertebrate conservation. These include a lack of public and political support, a lack of taxonomic or ecological knowledge, a need to refine survey methods and a need to understand species distributions. Some of this bias can also be explained by human preference for 'charismatic' species and by an understandable tendency to build on previous conservation success. However, the body of literature on butterfly conservation [REF], ecology [REF] and behaviour [REF] demonstrates the potential for invertebrate conservation. 

All of these reasons provide a compelling justification for undertaking conservation ecology research focusing on invertebrate species. 

Odonata, dragonflies and damselflies, are also relatively 'charismatic' invertebrate species and have received some considerable attention in the scientific literature [REF - model organisms book]. However, Odonata conservation is still in its relative infancy [REF] and even in countries with relatively few species [REF] where conservation is seen as a social priority [REF] such as the U.K. there is relatively little conservation action specifically for Odonata [REF].  The obstacles suggested by Cardoso _et al._ [REF] certainly apply to Odonata conservation [REF - Causnitzer] at both global and regional scales.  Indeed, priorities highlighted on the UK red list for Odonata include fully establishing species distributions and habitat requirements for several species [REF - red list], particularly those which inhabit remote areas and recommend efforts to monitor and understand population change. 

* Threats
The multi-stage life-cycle of odonata species leaves them vulnerable to threats in a number of ways [REF]. Indeed, around 25% of UK dragonfly and damselfly species are recognised in the red-data book [REF] The aquatic larval stage is vulnerable to aquatic pollution [REF], introduced species [REF] and land-use change [REF] and adult stages are particularly vulnerable to land-use change and habitat fragmentation [REF]. The largest proportions of both data deficient and threatened species are forest species which suggests that investigating the use and importance of forest habitats to dragonfly species is of particular importance. The majority of threatened dragonfly species are found in the tropics [REF] and species in Northern latitudes have been more comprehensively assessed [REF]. However, although conservation assessment may have taken place it does not necessarily follow that threats and habitat requirements are well known [REF]. Scale is also an important factor in assessing threats to individual populations. Globally, peatland dragonflies tend to be of lower conservation priority because of the large areas of peatland habitat available in the Northern Hemishpere (e.g. in Scandinavia) [REF]. However, peatland habitats can be locally threatened and are priority habitats in the UK [REF]. In these areas peatland species may be of conservation concern [REF].  


* WFD
* WFD Distribution
* WFD Habitat
White-faced darter (_Leucorhinnia dubia_) are a lowland bog specialist dragonfly species [REF]. In great Britain their distribution is heavily biased towards the Scottish highlands [REF]. There are several populations further south, in England, which have been the focus of conservation attention, including reintroduction programmes [REF].In Scotland their distribution stretches from Galloway in the South to Ross-shire in the North although the bulk of the population is found in the North on both sides of the Great Glen Fault. In the east they are found in parts of the Cairngorms and Grampians and they can be found on the West coast although they do not reach as far as the Inner Hebrides [REF].  The Scottish populations are thought to be declining but currently receive little conservation intervention [REF?]. The species is well monitored in some areas but their patchy distributions across a large landscape means that the species is thought to be under-recorded [REF]. White-faced darter habitat is well described in current ID guides and atlases although the evidence base for this is currently largely anecdotal or based on localised populations in England [REF]. As with many dragonfly species they have not been the subject of quantitative studies regarding habitat associations except at the very local scale [REF - BDS paper]. White-faced darter are generally found in acidic pools associated with pine or birch woodland [REF]. These pools may form part of pool complexes and it is not unusual for new sites to be located both close to current ones and at larger distances [REF]. They have a particularly strong association with Sphagnum moss [REF] and it is thought that well developed Sphagnum is an essential requirement for White-faced darter larval habitat [REF]. 

### Using SDMs and occupancy for conservation prioritisation and survey locations

Understanding species distributions is useful for the planning of future surveys [REF] and identifying priority areas for conservation action [REF]. As well as being useful for species which are poorly known [REF], or inhabit remote locations [REF] so their populations cannot be effectively monitored there are also species which are well know locally but poorly known over wider areas [REF]. Monitoring projects for various taxa have shown that biases can occur regarding the distribution of moinitoring effort or species records [REF]. This can be a particular problem for understudied taxa where expertise may be localised or restricted to only a few individuals [REF]. Where species are monitored sporadically [REF], through only incidental sightings [REF] or are extremely rare [REF] we might only have access to presence records for a species. Similarly, where taxa are monitored using volunteer recording schemes there is a trade-off between the information content of the data and the ease with which volunteers can complete records in the field [REF]. More complex surveys which result in information-rich data require more experienced/trained recorders and can limit the geographic coverage of the survey [REF]. On the other hand, simpler more widespread surveys can introduce issues with data quality due to inexperienced observers or patch records [REF]. Information-limited data restrict the methods available to investigate species distributions [REF] and can make it hard to project models onto unknown or under-recorded areas [REF]. In particular, this is a problem because, although species presence can be established, there are likely to be few or low quality records of species absence [REF]. 

Modelling methods to predict species distributions using data where absences are unavailable or uncertain requires the use of background values to differentiate suitable form unsuitable habitat [REF]. The most common of these algorithms, MaxEnt [REF], is a machine learning implementation of a point process regression which uses LASSO penalties to prevent overfitting [REF]. MaxEnt has been shown to outperform several other species distribution modelling methods [REF] and is widely used in conservation research. Here we use MaxEnt models to investigate the habitat requirements of White-faced Darter across the highlands of Scotland using records from the British Dragonfly Society as presence points along with environmental data reflecting tree cover and the presence of bogs. Understanding drivers of White-faced darter presence across large scales can contribute to conservation of this species by informing landscape management and will help to target future surveys for the species. 



<!--Can we add something on White faced darter distribution in Scotland? is there any paper done in this before or can we claim that we are the first to model the distribution in Scotland?. What about other dragonflies in Scotland and their conservation? I think there is enough on the issue of imperfect detection, but we have to put more biological and conservation background maybe -->


Methods
--------

##Data

<!------------- Records within 100m of each other were combined --------->

White-faced darter presence records between 2000 and 2015 were obtained from the British Dragonfly society. These records represent incidental records rather than a stratified recording scheme and can represent flying adults, larvae or exuvia. Additionally, the records related to White-faced darter refer only to presence of the species. We used spatial filtering to combine records within 100m of each other to reduce potential biases due to inaccurate georeferencing, individuals seen at distance and variability due to GPS accuracy [REF].

```{r}
## Read in data - including occupancy matrix
WFD.occ <- read.csv('WFDHighland_pres.csv')

### Remove any locations with only 1 record
WFD.occu.mat <- WFD.occ[,2:17]
WFD.occu.mat$length<-rowSums(!is.na(WFD.occu.mat[,1:16]))
WFD.occu.mat <-subset(WFD.occu.mat, WFD.occu.mat$length >1)
WFD.occu.mat <- WFD.occu.mat[1:17]

## Estimate WFD.detectability
WFD.UMF <- unmarkedFrameOccu(y = WFD.occu.mat[,2:17])
WFD.det <- occu(~1 ~1, data = WFD.UMF)
summary(WFD.det)
backTransform(WFD.det, 'det')
```



##Environmental variables
<!--------------- Matt to finish this section ------------>
Environmental variables used to predict occupancy of White-faced darter were downloaded from the EUNIS land-cover classification for Scotland (SOURCE AND LICENCE). The variables chosen were based on those described in the UK [REF] and European [REF] dragonfly atlases which specify wet, boggy areas within a woodland complex [REF]. We used established EUNIS classes [REF] to produce variables representing pineforest, coniferous forest, moorland, bogs, bog woodland and standing water. <!------------------- Need to check these variables -------------> Other potential variables such as mixed woodland or ancient woodland measures from other sources were not used in our analyses due to issues with collinearity. Instead we retained only the variables which we considered to have the strongest ecological connection with White-faced darter distribution based on the avialable literature [REF]. We converted our environmental variables to proportional maps per 1km at a 10m resolution from the original 10m resolution categorical landcover maps using the raster package [REF - Hijmans] in R [REF - CRAN]. These values were then aggregated to maps at a resultion of XXXX representing the average proportion of each variable from each 10m x 10m cell within the XXXm. Additionally we used data from the CEH CHESS dataset [REF] to represent potential Evapotranspiration (PET), indicating the 'wetness' of the habitat, which was resampled to match the resolution of our landcover data  using the raster package in R. 

```{r}
Env.vars <- stack(raster('EnvVars/bogs_50.img'), raster('EnvVars/bogwoodland_50.img'), raster('EnvVars/PET_emerge_50.img'), raster('EnvVars/conifer_50.img'), raster('EnvVars/moor_50.img'), raster('EnvVars/pineforest_50.img'), raster('EnvVars/standingwater_50.img'))
names(Env.vars) <- c("bog", "bogwoodland", "PET_emerge", "conifer", "moor", "pineforest", "standingwater")

### Check correlation between Env Vars
WFD.pres.env <- extract(Env.vars, WFD.pres[,1:2])
cor(WFD.pres.env)
### All < 0.5

```

# MaxEnt Model

Distribution models were fitted using MaxEnt version XXX through the dismo [REF] package in R. We used the ENMEvaluate package [REF] to find the 'best' combination of potential relationships with variables and the optimum regularisation parameter. We used Pearson's correaltion coefficient to check for correlation between predictor variables and in all cases these were less than 0.5. We tested linear, quadratic, product and threshold along with combinations of these. We used regularisation values between 0 and 2.5 at 0.5 increments. The optimum model was assessed by comparing AICcs across models [REF]. Once selected, we fitted the 'best' model using 10- crossvalidated replicates. Models were evaluated using the area under the reciever operating characterisitc curve (AUC; [REF - Fielding and Bell])as well as the True Skill Statistic [REF]. We used Moran's I to check for autocorrelation in the residuals (1 - predicted values at presence points). 

```{r}
### Models run on the server
## Candidate models testing 
### Create background points
#bg <- randomPoints(Env.vars[[1]], n=10000)
#bg <- as.data.frame(bg)
#WFD.eval.max <- ENMevaluate(WFD.pres[,1:2], Env.vars, bg, method='block', RMvalues=c(seq(1, 5, 0.5)),  fc=c('L','Q', 'LQ', 'P', 'LQP'), algorithm='maxent.jar', parallel  = TRUE)
#WFD.eval.max.1 <- ENMevaluate(WFD.pres[,1:2], Env.vars, bg, method='block', RMvalues=c(seq(1, 2.5, 0.5)),  fc=c('L'), algorithm='maxent.jar')
#save(WFD.eval.max.1, file = 'max.eval.results/WFD.max.eval.1.RData')
#rm(WFD.eval.max.1)
#gc()
#WFD.eval.max.2 <- ENMevaluate(WFD.pres[,1:2], Env.vars, bg, method='block', RMvalues=c(seq(2.5, 5, 0.5)),  fc=c('L'), algorithm='maxent.jar')
#save(WFD.eval.max.2, file = 'max.eval.results/WFD.max.eval.2.RData')
#rm(WFD.eval.max.2)
#gc()

#WFD.eval.max.3 <- ENMevaluate(WFD.pres[,1:2], Env.vars, bg, method='block', RMvalues=c(seq(1, 2.5, 0.5)),  fc=c('Q'), algorithm='maxent.jar')
#save(WFD.eval.max.3, file = 'max.eval.results/WFD.max.eval.3.RData')
#rm(WFD.eval.max.3)
#gc()

#WFD.eval.max.4 <- ENMevaluate(WFD.pres[,1:2], Env.vars, bg, method='block', RMvalues=c(seq(2.5, 5, 0.5)),  fc=c('Q'), algorithm='maxent.jar')
#save(WFD.eval.max.4, file = 'max.eval.results/WFD.max.eval.4.RData')
#rm(WFD.eval.max.4)
#gc()

#WFD.eval.max.5 <- ENMevaluate(WFD.pres[,1:2], Env.vars, bg, method='block', RMvalues=c(seq(1, 2.5, 0.5)),  fc=c('LQ'), algorithm='maxent.jar')
#save(WFD.eval.max.5, file = 'max.eval.results/WFD.max.eval.5.RData')
#rm(WFD.eval.max.5)
#gc()

#WFD.eval.max.6 <- ENMevaluate(WFD.pres[,1:2], Env.vars, bg, method='block', RMvalues=c(seq(2.5, 5, 0.5)),  fc=c('LQ'), algorithm='maxent.jar')
#save(WFD.eval.max.6, file = 'max.eval.results/WFD.max.eval.6.RData')
#rm(WFD.eval.max.6)
#gc()

#WFD.eval.max.7 <- ENMevaluate(WFD.pres[,1:2], Env.vars, bg, method='block', RMvalues=c(seq(1, 2.5, 0.5)),  fc=c('P'), algorithm='maxent.jar')
#save(WFD.eval.max.7, file = 'max.eval.results/WFD.max.eval.7.RData')
#rm(WFD.eval.max.7)
#gc()

#WFD.eval.max.8 <- ENMevaluate(WFD.pres[,1:2], Env.vars, bg, method='block', RMvalues=c(seq(2.5, 5, 0.5)),  fc=c('P'), algorithm='maxent.jar')
#save(WFD.eval.max.8, file = 'max.eval.results/WFD.max.eval.8.RData')
#rm(WFD.eval.max.8)
#gc()

#WFD.eval.max.9 <- ENMevaluate(WFD.pres[,1:2], Env.vars, bg, method='block', RMvalues=c(seq(1, 2.5, 0.5)),  fc=c('LQP'), algorithm='maxent.jar')
#save(WFD.eval.max.9, file = 'max.eval.results/WFD.max.eval.9.RData')
#rm(WFD.eval.max.9)
#gc()

#WFD.eval.max.10 <- ENMevaluate(WFD.pres[,1:2], Env.vars, bg, method='block', RMvalues=c(seq(2.5, 5, 0.5)),  fc=c('LQP'), algorithm='maxent.jar')
#save(WFD.eval.max.10, file = 'max.eval.results/WFD.max.eval.10.RData')
#rm(WFD.eval.max.10)
#gc()
```

```{r}
load('max.eval.results/WFD.max.eval.1.RData')
load('max.eval.results/WFD.max.eval.2.RData')
load('max.eval.results/WFD.max.eval.3.RData')
load('max.eval.results/WFD.max.eval.4.RData')
load('max.eval.results/WFD.max.eval.5.RData')
load('max.eval.results/WFD.max.eval.6.RData')
load('max.eval.results/WFD.max.eval.7.RData')
load('max.eval.results/WFD.max.eval.8.RData')
load('max.eval.results/WFD.max.eval.9.RData')
load('max.eval.results/WFD.max.eval.10.RData')

max.eval.results.full <- rbind(WFD.eval.max.1@results, WFD.eval.max.2@results, WFD.eval.max.3@results, WFD.eval.max.4@results, WFD.eval.max.5@results, WFD.eval.max.6@results, WFD.eval.max.7@results, WFD.eval.max.8@results, WFD.eval.max.9@results, WFD.eval.max.10@results)

max.eval.results.full$newDelta <- max.eval.results.full$AICc - min(max.eval.results.full$AICc)

max.eval.results.full[which(max.eval.results.full$newDelta <= 2),]

<<<<<<< HEAD
# Scale and centre siteCovs
siteCovs(WFD.UMF) <- scale(siteCovs(WFD.UMF))
sC.scale <- scale(siteCovs(WFD.UMF))
obsCovs(WFD.UMF) <- scale(obsCovs(WFD.UMF))
oC.scale <- scale(obsCovs(WFD.UMF))
```

##Occupancy model

<!------------------ ACHAZ - explain the model etc. here ---------------->

<!------------- We don't currently try any observation covariates - I know these didn't help but I'm thinking we should test that here, too? I'm not sure our model selection was the most methodical in the end. 
AvH 13/08: Try a dynamic model dividing in 5 years slots? 
-------------->
<!--------------------- Adapted the model selection below to follow the Kery book (Applied Heirarchical modelling). Separate model selection step for detectability but not followed the stepwise approach to predictors. ------>

```{r occupancy}
### Candidate models obsCovs
WFD.occu.obs.1 <- occu(~1 ~1, WFD.UMF)
WFD.occu.obs.2 <- occu(~AWI ~1, WFD.UMF)
WFD.occu.obs.3 <- occu(~wildness ~1, WFD.UMF)
WFD.occu.obs.4 <- occu(~wind ~1, WFD.UMF)
WFD.occu.obs.5 <- occu(~tas ~1, WFD.UMF)
#WFD.occu.obs.6 <- occu(~AWI + wildness + wind + tas ~1, WFD.UMF)
WFD.occu.obs.7 <- occu(~AWI + wildness + wind ~1, WFD.UMF)
WFD.occu.obs.8 <- occu(~AWI + wildness + tas ~1, WFD.UMF)
WFD.occu.obs.9 <- occu(~AWI + wildness + tas ~1, WFD.UMF)
WFD.occu.obs.10 <- occu(~AWI + wind + tas ~1, WFD.UMF)
WFD.occu.obs.11 <- occu(~wildness + wind + tas ~1, WFD.UMF)
WFD.occu.obs.12 <- occu(~AWI + wildness ~1, WFD.UMF)
WFD.occu.obs.13 <- occu(~AWI + wind ~1, WFD.UMF)
WFD.occu.obs.14 <- occu(~wildness + wind ~1, WFD.UMF)
WFD.occu.obs.15 <- occu(~AWI + tas ~1, WFD.UMF)
WFD.occu.obs.16 <- occu(~wildness + tas ~1, WFD.UMF)
#WFD.occu.obs.17 <- occu(~wind + tas ~1, WFD.UMF)

WFD.obs.fits <- fitList(WFD.occu.obs.1, WFD.occu.obs.2, WFD.occu.obs.3, WFD.occu.obs.4, WFD.occu.obs.5, WFD.occu.obs.7, WFD.occu.obs.8, WFD.occu.obs.9, WFD.occu.obs.10, WFD.occu.obs.11, WFD.occu.obs.12, WFD.occu.obs.13, WFD.occu.obs.14, WFD.occu.obs.15, WFD.occu.obs.16)
WFD.modSel.obs <- modSel(WFD.obs.fits)
WFD.modSel.obs



### Candidate models siteCovs
WFD.occu.1 <- occu(~wildness + wind + tas ~1, WFD.UMF)
WFD.occu.2 <- occu(~wildness + wind + tas~ pineforest*wet + pet_emerge, WFD.UMF)
WFD.occu.3 <- occu(~wildness + wind + tas ~pineforest + pet_emerge + bogs, WFD.UMF)
WFD.occu.4 <- occu(~wildness + wind + tas ~pineforest + pet_emerge + wet, WFD.UMF)
WFD.occu.5 <- occu(~wildness + wind + tas ~pineforest + bogs, WFD.UMF)
WFD.occu.6 <- occu(~wildness + wind + tas ~pineforest + wet, WFD.UMF)
WFD.occu.7 <- occu(~wildness + wind + tas ~pineforest + pet_emerge, WFD.UMF)
WFD.occu.8 <- occu(~wildness + wind + tas ~pineforest*pet_emerge + wet, WFD.UMF)
WFD.occu.9 <- occu(~wildness + wind + tas ~pineforest*pet_emerge + bogs, WFD.UMF)
WFD.occu.10 <- occu(~wildness + wind + tas ~pineforest*pet_emerge, WFD.UMF)
WFD.occu.11 <- occu(~wildness + wind + tas ~pineforest*wet, WFD.UMF)
WFD.occu.12 <- occu(~wildness + wind + tas ~pineforest*bogs, WFD.UMF)
WFD.occu.13 <- occu(~wildness + wind + tas ~pineforest, WFD.UMF)
WFD.occu.14 <- occu(~wildness + wind + tas ~coniferpine + pet_emerge + wet, WFD.UMF)
WFD.occu.15 <- occu(~wildness + wind + tas ~coniferpine*wet + pet_emerge, WFD.UMF)
WFD.occu.16 <- occu(~wildness + wind + tas ~coniferpine*bogs + pet_emerge, WFD.UMF)

WFD.site.fits <- fitList(WFD.occu.1, WFD.occu.2, WFD.occu.3, WFD.occu.4, WFD.occu.5, WFD.occu.6, WFD.occu.7, WFD.occu.8, WFD.occu.9, WFD.occu.10, WFD.occu.11, WFD.occu.12, WFD.occu.13, WFD.occu.14, WFD.occu.15, WFD.occu.16)
WFD.modSel <- modSel(WFD.site.fits)
WFD.modSel

WFD.occu.full <- occu(~AWI+wildness+wind+tas ~coniferpine*pineforest*bogs*wet + pet_emerge, WFD.UMF)

d1 <- dredge(WFD.occu.full, rank = AIC)

  #### Quadratic for tree variable or 'wet' variable does not improve model
=======
## Fit MaxEnt model using dismo
WFD.max <- maxent(x = Env.vars, p = WFD.train[,1:2], args=c('replicates = 10', 'replicatetype = crossvalidate', 'betamultiplier=1', 'pictures=true', 'linear=true', 'quadratic=true', 'product=false', 'threshold=false', 'hinge=false', 'responsecurves=true', 'jackknife = true'))
>>>>>>> 16b6bf280b573eb658e8971e4d8fbacb67a81b80
```

Results
--------

## Model

Model selectio indicated that the 'best' combination of parameters for our model was to use only linear and quadratic features with a beta multiplier of one (Table ???). This model showed a good fit to the data (AUC = 0.87, TSS = ??). The most important variable in the model (69% variable contribution) was the percentage of pineforest (Figure ??). Other variables were considerably less important with the percentage of standing water (9.6%), percentage of bog woodland (8%) and potential evapotranspiration during emergence (7%) having slightly higher contributions than percentage of bogs, percentage of conifer and percentage of moorland which all have percentage contributions of less than 5%.  

Response curves indicate that the dependence of suitability on the percentage of pineforest shows a non-linear response (Figure ???). Although generally higher proportions of forest increase the relative suitability value, the maximum value is found at around 60% forest cover. Similarly, the highest relative suitability is predicted when the proportion of bog woodland is around 35% and the proportion of standing water around 30%.

Projecting the model onto environmental data for scotland produces maps which accurately predict many current strongholds for White-faced darter. In particular, the maps sucessfully predict Abernethy forest and Glen Affric. However, these maps fail to predict a well known population of White-faced darter at Monadh Mor on the Black Isle. The model also predicts several areas in the Grampians as suitable where we have no current records of White-faced darter. 

<!-------What else do we need in here? ---------------------->

## Predicted distribution




Discussion
------------

## Model results

Our models suggest that the strongest drivers of White-faced darter distribution are increasing proportions of pine forest along with bog woodland and standing water. These results provide quantitative support for the current descriptions of White-faced darter habitat [REF] and also agree with local-scale descriptions of habitat associations [REF]. Our maps of predicted distribution predict strongly in several core areas for White-faced darter such as Abernethy forest and Glen Affric. However, in other areas our predictions are less successful. In fact, we miss one well know location entirely and are unable to successfully predict the population found at Monadh Mor on the Black Isle. Our models also only consider pine forest and do not include the birch forests suggested by several sources [REF]. We see this as a limitation of our model rather than an indication that these are not important [REF]. <!-- Did we not try to include birch forest or did it just not work? in any case we have to explain why this is a limitation of our model --> More successfully, our models predict a high probability of White-faced darter presence in areas which the species was not found until very recently, locations not included in our data (e.g. <!----- Check darter magazine and maps - not sure this is true any more ------->.

## Areas for survey or protection
Our models currently predict several areas where White-faced darter have not been recorded yet. This suggests that White-faced darter may be under-recorded across the highlands. Ideally efforts could be made to survey for this species in remote areas in the central highlands, grampians and at the species northern range limit in Sutherland. Remote areas are often difficult to obtain data for when relying on volunteer recorders. One approach to this in UK bird surveys is to encourage volunteers from elsewhere in the country to adopt remote squares [REF]. This method may be particularly applicable to dragonfly recorders who are fewer in number than those who contribute bird records [REF]. Two larger areas of habitat are predicted by our model based around Abernethy forest and Glen Affric. These two sites are well known locations for the species and our models suggest they may be important sites for the conservation of the species. Large areas of contiguous habitat are likely to hold larger populations. White-faced darter should continue to be considered in management plans for these two sites with the maintenance of bog pool systems within woodland as a focus. 


## Model evaluation
Data from volunteer surveys is challenging to analyse because of the potential biases which may be present. Records are often biased spatially with more records close to population centres. However, in this region, where population density is low, this does not appear to be the case. However, records in this region may be influenced by the ease of access and be biased towards sites close to roads or those which are well known. This represents a limitation with our approach and, ideally, our predictions could be used for a wider, more comprehensive survey of this species which expands the survey coverage across the highlands. Our approach was also limited by the scale of habitat data available. Although very high resolution data are available, over large areas the volume of data involved in these models is prohibitive. As such, data were aggregated to larger areas and so predictions are likely to be considerably coarser than a dragonfly's perception of habitat features. 

## Conservation implications for WFD

Our models are in agreement with the current management recommendations for this species [REF] and emphasise the importance of lowland peat pool complexes within woodland. Our models particularly identify pinewood and suggest that native pinewoods may be particularly important although current management practices also suggest that some broadleaved species would be equally beneficial [REF]. Indeed these species are usually those found in edge habitats as part of native Caledonian pinewood [REF]. In terms of management, our models suggest that the maintenance of pool complexes within woodland is the most important objective. Our models predict that where woodland cover is high, the proportion of 'wet' habitat can be relatively small but without the forest cover even large areas of 'wet' habitat may not be suitable. This also suggest that White-faced darter are likely to be impacted by drainage as part of woodland management and this should eb avoided where this species is present. This aslo adds support to current management operations as part of the White-faced darter reintroduction project in Delamere forest, England where areas are being re-wetted to provide more suitable habitat for the reintroduced population [REF - Meredith BDS].

## Potential future work
other odonata

Our models currently only use presence records to investigate distribution. However, records of species not only depend on the distribution of that species but also on the detectability of the species [REF]. Presence-only methods can risk modelling the detectability of a species rather than its true distribution [REF]. We are confident that this is not the case in our model because these presence records are part of a wider dataset of dragonfly records where the presence of other species indicates some considerable searching by dedicated recorders. However, in future we recommend that data are collected in a way which facilitates the incorporation of detectability into models of dragonfly distributions. This would involve making repeat visits to sites [REF] and indicating when a list of all species present has been provided to the national recording scheme [REF]. 

Wide-scale monitoring and recording of odonata is still in its infancy compared to other taxa. Currently the distributions of many species, even in well recorded parts of the world, are known geographically but the ecological drivers of those distributions are less well understood. Our model demonstrates that investigating the distributions is possible even with current monitoring data. Indeed a useful next step would be to extend this modelling approach to include further species of conservation concern. The Scottish highlands is home to several other Odonata species which, along with White-faced darter, are under-recorded and potentially threatened by land use change [REF]. Adapting the models used in our analysis to include multiple species could benefit conservation planning for Odonata by identifying 'hotspots' for these specialists. 

Our models used all records of White-faced darter which included larval records, exuvia counts and adult sightings. These three different record types are likely to have different detectabilities [REF] and, in fact, represent different levels of information about White-faced darter populations. Not all locations where adults are seen are suitable for the full reproductive cycle of Odonata [REF] and records fo larvae or exuvia are required to truly demonstrate successful breeding [REF]. An extension of our method which could sue this information to make more accurate estimates of White-faced darter distributions would be to use mutli-stage models of occupancy [REF] which could use adult presence and confirmed breeding as separate outcomes within the model. However, current recording of Odonata within the Scottish highlands does not represent enough information to fully parameterise such a model. For this to be successful we would need to increase the number of repeat visits to sites and particular increase the number of repeats which sought larvae or exuvia rather than just adults [REF]. 



# COnnectivity and population structure




## Conclusions
